Temporal distribution of roadkills
========================================================
Roadkills are a frequent outcome of interaction of transportation and wildlife. Research shows, that animals follow certain cycles, and animal-vehicle crashes occur often around sunrise and sunset ([Haikonen and Summala, 2001](http://www.sciencedirect.com/science/article/pii/S074937970100352X)).

This document shows nice way, how to visualize such data.

```{r load.packages, include=FALSE, cache=FALSE}
library(maptools)
library(plotrix)
```

There are 30 000 roadkills available, each has recorded position, date and time.

```{r load.prepare.data}
# read data
roadkills <- read.csv('roadkills.csv')
str(roadkills)
# transform date and time
datetime <- as.POSIXct(roadkills$datetime)

# calculate time as float
hours <- as.POSIXlt(datetime)$hour 
minutes <- as.POSIXlt(datetime)$min
time <- hours + minutes/60
```

We can calculate time of sunrise and sunset for every accident, based on its locations (functions from package **maptools**).

```{r calc.sunrise.sunset}
# calculate sunrise and sunset for every colision
sunrise <- sunriset(as.matrix(roadkills[,2:3]), datetime, direction='sunrise')*24
sunset <- sunriset(as.matrix(roadkills[,2:3]), datetime, direction='sunset')*24
```

Then, we will calculate the time of sunset and sunrise through the year.

```{r sunset.through.year}
# culculate time of sunrise and sunset through the year
center <- matrix(c(15.33, 49.75), ncol=2) # center of cyech republic as a reference points
seq_datum <- seq.Date(from=as.Date("2010/1/1"), to=as.Date("2010/12/31"),by='day') # sequence of days
sun <- c()
for (i in 1:length(seq_datum)) {
  d <- seq_datum[i]
  sr <- sunriset(center,as.POSIXct(d), direction='sunrise')*24
  ss <- sunriset(center,as.POSIXct(d), direction='sunset')*24
  sun <- rbind(sun,c(sr, ss))
}
```

And finally, plot numbers of accidents for every month and hour.

```{r plot, fig.height=6, fig.width=8}
# calculate number of accidents by month and hours
seq1 <- c(0, 31, 59, 90, 120, 151, 181, 212, 243, 273, 304, 334, 365)
seq2 <- seq(0, 24, 1)

# day of year
doy <- as.numeric(format(datetime, '%j'))

counts <- c()
for(i in 1:(length(seq1)-1)) {
  for(j in 1:(length(seq2)-1)){
    limit_x <- c(seq1[i], seq1[i+1])
    limit_y <- c(seq2[j], seq2[j+1])
    n <- sum(doy>limit_x[1] & doy<=limit_x[2] & hours>=limit_y[1] & hours<limit_y[2])
    counts <- c(counts,n)
  }  
}

# prepate empty plot with axes
par(xpd=NA)
plot(as.numeric(format(datetime, '%j')), time, pch=16, cex=.4, xlim=c(0,460), ylim=c(0,24), axes=F, xlab='Month', ylab='Hour', type='n')
axis(side=1, at=seq(15, 365, 30), labels=c('J', 'F', 'M', 'A', 'M', 'J', 'J', 'A', 'S', 'O', 'N', 'D'), , cex.axis=.6)
axis(side=2, at=seq(0,24,2), labels=seq(0,24,2), cex.axis=.6)

# color palette from yellow to red
q <- quantile(range(counts),seq(0,1,.1))
colors <- color.gradient(c(1,1), c(1,0), c(0,0), nslices=10)

# plot counts
counter <- 1
for(i in 1:(length(seq1)-1)) {
  for(j in 1:(length(seq2)-1)){
    limit_x <- c(seq1[i], seq1[i+1])
    limit_y <- c(seq2[j], seq2[j+1])
    val <- counts[counter]
    
    counter2 <- 2    
    while(val>q[counter2]){
      counter2 <- counter2+1
    }
    
    col <- colors[counter2-1]
    
    polygon( x=c(limit_x[1], limit_x[2], limit_x[2], limit_x[1]), y=c(limit_y[1], limit_y[1], limit_y[2], limit_y[2]), border=NA, col=col)
    
    counter <- counter+1
  }  
}

# add lines of sunrise and sunset
lines(1:365, sun[,1], lwd=2)
lines(1:365, sun[,2], lwd=2)
title('Daily and seasonal distribution of roadkills\nOctober 2006 - December 2012')

# add legend
counter <- 1
seq3 <- seq(0,24, length.out=(length(colors)+1))
for(i in 1:(length(seq2)-1)) {
  limit_x <- c(400,430)
  limit_y <- c(seq3[i], seq3[i+1])
  col <- colors[counter]
  polygon( x=c(limit_x[1], limit_x[2], limit_x[2], limit_x[1]), y=c(limit_y[1], limit_y[1], limit_y[2], limit_y[2]), border=NA, col=col)
  counter <- counter+1
}

text(rep(445,(length(colors)+1)), seq3+0.5, adj=c(0,1),
     labels=c('0 % = 18', '10 %', '20 %', '30 %', '40 %', '50 %', '60 %', '70 %', '80 %', '90 %', '100 % = 517'),
     cex=.6,)
axis(1, at=430, labels='Nr. of\nAccidents', tick=F, cex.axis=.8)
```