Calculate Stationing of points on line
========================================================

R currently does not include functions to create [Linear referencing system](http://en.wikipedia.org/wiki/Linear_referencing), but it is rather straightforward to create one. This document will show functions to do so.

```{r load, message=FALSE}
# load packages
library(maptools)
```

First, load sample points and lines.

```{r fig.width=10, fig.height=10}
# load points
p <- readShapePoints('data/points')
# load lines
l <- readShapeLines('data/lines')

# plot
plot(l)
plot(p, add=T, cex=1, pch=16, col='grey60')
```

The process of calculation of stationing includes two steps: snapping points to lines amd then calculating the position of points on the line.

First: a function, which determines, if point lies on line between two another points needs to be defined for further use.

```{r point.on.line}
pointOnLine <- function(point,line_start,line_end){ # each of input parameters is pair of coordinates [x,y]
  if(identical(point,line_start) | identical(point,line_end)){return(TRUE)} # check, if the points cooincides with start/end point
  if(point[1]>max(c(line_start[1],line_end[1])) | point[1]<min(c(line_start[1],line_end[1])) | point[2]>max(c(line_start[2],line_end[2])) | point[2]<min(c(line_start[2],line_end[2]))){
    return(FALSE) # if the point is out of the bounding box of the line, return false
  }
  if(line_start[2]==line_end[2]){
    slope<-0
  } else if(line_start[1]==line_end[1]){
    return(T)
  } else {
    slope <- (line_start[2]-line_end[2])/(line_start[1]-line_end[1])
  }
  intercept <- -slope*line_start[1] + line_start[2]
  onLine <- round(point[2],digits=0)==round((slope*point[1]+intercept),digits=0)
  return(onLine)
}
```

Than, a function, which calculates the stationing:

```{r calculate.stationing}
calculateStationing <- function(points, lines, maxDist=NA){
  
  # snap points to lines from package maptools
  snapped <- snapPointsToLines(points, lines, maxDist, withAttrs=TRUE)
  
  
  stationing <- c()
  
  for (i in 1:length(snapped)){
    crds_p <- coordinates(snapped[i,])
    line <- lines[snapped[i,]$nearest_line_id,]
    crds_l <- coordinates(line)[[1]][[1]]
    d <- 0
    for(j in 2:nrow(crds_l)){
      onLine <- pointOnLine(point=crds_p,line_start=crds_l[j-1,],line_end=crds_l[j,])
      if(onLine){
        d0 <- sqrt((crds_p[1]-crds_l[j-1,1])^2 + (crds_p[2]-crds_l[j-1,2])^2)
        stationing <- c(stationing,round(d+d0))
        break
      }
      d <- d + sqrt((crds_l[j,1]-crds_l[j-1,1])^2 + (crds_l[j,2]-crds_l[j-1,2])^2)
    }
  }
  
  snapped$stationing <- stationing
  return(snapped)
}
```

However, function `snapPointsToLines` includes a (possible) bug, so it needs to be redefined:

```{r}
snapPointsToLines <- function (points, lines, maxDist = NA, withAttrs = TRUE) 
{
    require("rgeos")
    if (is(points, "SpatialPoints") && missing(withAttrs)) 
        withAttrs = FALSE
    if (!is.na(maxDist)) {
        w = gWithinDistance(points, lines, dist = maxDist, byid = TRUE)
        validPoints = apply(w, 2, any)
        validLines = apply(w, 1, any)
        points = points[validPoints, ]
        lines = lines[validLines, ]
    }
    d = gDistance(points, lines, byid = TRUE)
    nearest_line_index = apply(d, 2, which.min)
    coordsLines = coordinates(lines)
    coordsPoints = coordinates(points)
    mNewCoords = vapply(1:length(points), function(x) nearestPointOnLine(coordsLines[[nearest_line_index[x]]][[1]], 
        coordsPoints[x, ]), FUN.VALUE = c(0, 0))
    if (!is.na(maxDist)) 
        nearest_line_id = as.numeric(rownames(d)[nearest_line_index]) + 
            1
    else nearest_line_id = nearest_line_index
    if (withAttrs) 
        df = cbind(points@data, nearest_line_id)
    else df = data.frame(nearest_line_id, row.names = names(nearest_line_index))
    SpatialPointsDataFrame(coords = t(mNewCoords), data = df, 
        proj4string = CRS(proj4string(points)))
}
```

And finally we can calculate the stationing of real data:

```{r calculate.stationing2}
#calculate stationing
snapped.p <- calculateStationing(p,l)
head(snapped.p@data)
```

For better vizualization, we can also connect original points with their snapped self. (**Note:** This works only if no `maxDist` was set)

```{r connector.lines, fig.width=10, fig.height=10}
# create connector lines
connector.lines <- list()
for (i in 1:length(p)){
  connector.lines[[i]] <- Lines(slinelist=list(Line(coords=rbind(coordinates(p[i,]), coordinates(snapped.p[i,])))) ,ID=as.character(i))
}
connector.lines <- SpatialLines(connector.lines)

# plot zoomed out area
plot(l, axes=T, xlim=c(-515000, -510000), ylim=c(-1188000, -1186000))
plot(p, pch=16, add=T)
plot(snapped.p, pch=16, add=T, col='red')
plot(connector.lines, add=T, col='grey30', lty=3)
```