*k*-Shortest Paths
========================================================

This document will show, how to find *k*-shortest paths in a graph using [igraph](http://igraph.sourceforge.net/) library. Typical example can be finding route between two places and wanting to have several alternatives to choose from.

The procedure will be similar to the one used by [Yen's algorithm](http://en.wikipedia.org/wiki/Yen%27s_algorithm). The idea is following: 
+ Assume, we have shortest path between two nodes in a graph
+ Iterate through edges of shortest path, removing one of them each iteration, and calculating shortest path - *alternative routes*
+ Shortest of *alternative routes* is second shortest path
+ Shortest path *n* will be shortest alternaive route to shortest paths *n-1, n-2, ..., 1*

Implementation in R is following:

```{r functions}
# find k shortest paths
k.shortest.paths <- function(graph, from, to, k){
  # first shortest path
  k0 <- get.shortest.paths(graph,from,to, output='both')
  
  # number of currently found shortest paths
  kk <- 1
  
  # list of alternatives
  variants <- list()
  
  # shortest variants
  shortest.variants <- list(list(g=graph, path=k0$epath, vert=k0$vpath, dist=shortest.paths(graph,from,to)))
  
  # until k shortest paths are found
  while(kk<k){
    # take last found shortest path
    last.variant <- shortest.variants[[length(shortest.variants)]]              
        
    # calculate all alternatives
    variants <- calculate.variants(variants, last.variant, from, to)
        
    # find shortest alternative
    sp <- select.shortest.path(variants)
    
    # add to list, increase kk, remove shortest path from list of alternatives
    shortest.variants[[length(shortest.variants)+1]] <- list(g=variants[[sp]]$g, path=variants[[sp]]$variants$path, vert=variants[[sp]]$variants$vert, dist=variants[[sp]]$variants$dist)
    kk <- kk+1
    variants <- variants[-sp]
  }
  
  return(shortest.variants)
}

# found all alternative routes
calculate.variants <- function(variants, variant, from, to){
  # take graph from current path
  g <- variant$g
  
  # iterate through edges, removing one each iterations
  for (j in unlist(variant$path)){
    newgraph <- delete.edges(g, j) # remove adge
    sp <- get.shortest.paths(newgraph,from,to, output='both') # calculate shortest path
    spd <- shortest.paths(newgraph,from,to) # calculate length
    if (spd != Inf){ # the the path is found
      if (!contains.path(variants, sp$vpath)) # add to list, unless it already contains the same path
      {
        variants[[length(variants)+1]] <- list(g=newgraph, variants=list(path=sp$epath, vert=sp$vpath, dist=spd))
      }
    }
  }
  
  return(variants)
}

# does a list contain this path?
contains.path <- function(variants, variant){
  return( any( unlist( lapply( variants, function(x){ identical(x$variant$vert,variant) } ) ) ) )
}

# which path from the list is the shortest?
select.shortest.path <- function(variants){
  return( which.min( unlist( lapply( variants, function(x){x$variants$dist} ) ) ) )
}
```

And lets see some examples: 

(warnings tell me, that removing some edges leads to graph, where the two nodes can not be reached)

```{r example}
library(igraph)
g <- read.graph('graph.gml', format='graphml')

k.shortest.paths(g, from=2795, to=2839, k=5)
```

Resulting object is a list, each item of the list includes graph, indexes of vertices, indexes of edges and distance. Vertices are the same in every graph, and hence their indexes as well. Edges are different, and therefore indexes of edges can differ.
